---
title: "ggplot: Part III"
subtitle: "Workshop on Plotting in R"
author: "`r paste0('<b>Lokesh Mano</b> • ',format(Sys.time(), '%d-%b-%Y'))`"
---

```{r, include=FALSE}
hooks = knitr::knit_hooks$get()
hook_foldable = function(type) {
  force(type)
  function(x, options) {
    res = hooks[[type]](x, options)
    
    if (isFALSE(options[[paste0("fold.", type)]])) return(res)
    
    paste0(
      "<details><summary>", type, "</summary>\n\n",
      res,
      "\n\n</details>"
    )
  }
}
knitr::knit_hooks$set(
  output = hook_foldable("output"),
  plot = hook_foldable("plot")
)
```

```{r,child="assets/header-lab.Rmd"}
```


```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(dplyr)
library(ggplot2)
library(reshape2)
data("iris")
gc <- read.table("data/counts_raw.txt", header = T, row.names = 1, sep = "\t")
md <- read.table("data/metadata.csv", header = T, sep = ";")
rownames(md) <- md$Sample_ID

library(tidyverse)
gc_long <- gc %>%
  rownames_to_column(var = "Gene") %>%
  gather(Sample_ID, count, -Gene) %>% 
  full_join(md, by = "Sample_ID") %>% 
  select(Sample_ID, everything()) %>% 
  select(-c(Gene,count), c(Gene,count)) 
gc_long$Sample_Name <- factor(gc_long$Sample_Name, levels = c("t0_A","t0_B","t0_C","t2_A","t2_B","t2_C","t6_A","t6_B","t6_C","t24_A","t24_B","t24_C"))
gc_long$Time <- factor(gc_long$Time, levels = c("t0","t2","t6","t24"))
gc_long$Replicate <- factor(gc_long$Replicate, levels = c("A","B","C"))

se <- function(x) sqrt(var(x)/length(x))
```

Now that we have covered the important aspects of `ggplot`, meaning getting the actual plot you wanted, let us now look into secondary elements of the plot.

# Modifying Axes

## Breaks

If we look at the `iris` data plot that we made before: 

```{r, fold.plot=FALSE}
ggplot(data=iris,mapping=aes(x=Petal.Length,y=Petal.Width))+
  geom_point(aes(color=Sepal.Width))+
  geom_smooth(method="lm")
```

The continuous x axis breaks are with 2,4,6 and so on. If we would like to have 1,2,3… We change this using `scale_x_continuous()` and `breaks`.

```{r}
ggplot(data=iris,mapping=aes(x=Petal.Length,y=Petal.Width))+
  geom_point(aes(size=Sepal.Width, color=Species))+
  geom_smooth(method="lm") +
  scale_x_continuous(breaks = 1:7)
```

You can do the same with `y-axis`. 

```{r}
ggplot(data=iris,mapping=aes(x=Petal.Length,y=Petal.Width))+
  geom_point(aes(size=Sepal.Width, color=Species))+
  geom_smooth(method="lm") +
  scale_x_continuous(breaks = 1:7) +
  scale_y_continuous(breaks = seq(0,3,0.5))
```

## Limits

By using `limits`, we can also decide on the parts to plot to be shown:

```{r}
ggplot(data=iris,mapping=aes(x=Petal.Length,y=Petal.Width))+
  geom_point(aes(size=Sepal.Width, color=Species)) +
  geom_smooth(method="lm") +
  scale_x_continuous(limits=c(3, 7))
```

We can do the same with discrete x values like in the case of our `gene counts` dataset.

```{r}
gc_long %>% 
  group_by(Time, Replicate) %>% 
  summarise(mean=mean(log10(count +1)),se=se(log10(count +1))) %>%
  ggplot(aes(x=Time, y=mean, fill = Replicate)) + 
  geom_col() + 
  scale_x_discrete(limits=c("t0","t24"))
```

One can also use `xlim()` and `ylim()` functions that function the same as `limits` with `scale_x_continous()` or `scale_x_discrete()` 

## Names 

You can also customize the axis labels using the `name` option within `scale_x_continous` and `scale_y_continous`.

```{r, warning=FALSE, message=FALSE}
ggplot(data=iris,mapping=aes(x=Petal.Length,y=Petal.Width))+
  geom_point(aes(size=Sepal.Width, color=Species))+
  geom_smooth(method="lm") +
  scale_x_continuous(name = "Length", breaks = 1:7) +
  scale_y_continuous(name = "Width", breaks = seq(0,3,0.5))
```

with `labels` in combination with the `scales` package, one can change or make the unit of the axis look more comprehensible, when needed. Like using `percentage` option or `scientific` option.

```{r, warning=FALSE, message=FALSE}
library(scales)
ggplot(data=iris,mapping=aes(x=Petal.Length,y=Petal.Width))+
  geom_point(aes(size=Sepal.Width, color=Species))+
  geom_smooth(method="lm") +
  scale_y_continuous(name = "Width", breaks = seq(0,3,0.5), labels = scientific)
```

# Legends

There are many ways to control the legends, below are some of the examples:
First by using `guides()` function.

```{r}
ggplot(data=iris,mapping=aes(x=Petal.Length,y=Petal.Width))+
  geom_point(aes(color=Species,size=Sepal.Width))+
  guides(size="none")
```

We can also turn off legends by geom.

```{r, warning=FALSE, message=FALSE}

ggplot(data=iris,mapping=aes(x=Petal.Length,y=Petal.Width))+
  geom_point(aes(color=Species,size=Sepal.Width),show.legend=FALSE)
```

## Editing legends

The legends can be edited by `scale_<aesthetic>_<discrete or continous>` function that we have been using. Take the below figure for example, we have the `Sepal.Width` and the `Species` with the `size` and `color` aestheitcs respectively.

```{r, fold.plot=FALSE, warning=FALSE, message=FALSE}
ggplot(data=iris,mapping=aes(x=Petal.Length,y=Petal.Width))+
  geom_point(aes(size=Sepal.Width, color=Species))+
  geom_smooth(method="lm")
```

Let's try to edit the legends here like mentioned before:

```{r, fold.plot=FALSE, warning=FALSE, message=FALSE}
ggplot(data=iris,mapping=aes(x=Petal.Length,y=Petal.Width))+
  geom_point(aes(size=Sepal.Width, color=Species))+
  geom_smooth(method="lm") +
  scale_size_continuous(name = "Width of Sepal") +
  scale_color_discrete(name = "Organism", labels = c("Iris setosa", "Iris versicolor", "Iris virginica"))
```

## Moving legends

Legends can be moved around using theme.

```{r, warning=FALSE, message=FALSE}
ggplot(data=iris,mapping=aes(x=Petal.Length,y=Petal.Width))+
  geom_point(aes(color=Species,size=Sepal.Width)) +
  theme(legend.position="top",
        legend.justification="right")
```

Legend rows can be controlled in a finer manner.

```{r, warning=FALSE, message=FALSE}
ggplot(data=iris,mapping=aes(x=Petal.Length,y=Petal.Width))+
  geom_point(aes(color=Species,size=Sepal.Width))+
  guides(size=guide_legend(nrow=2,byrow=TRUE),
         color=guide_legend(nrow=3,byrow=T))+
  theme(legend.position="top",
        legend.justification="right")
```

# Themes

Now that we started into `theme()`, it is possible to much more editing of the plot with this function. Let us look into some of the parameters that would be very helpful to work with. 

## Axis style

You can change the style of the axis texts in the following way:

```{r, warning=FALSE, message=FALSE}
ggplot(data=iris,mapping=aes(x=Petal.Length,y=Petal.Width)) +
  geom_point(aes(color=Species,size=Sepal.Width)) +
  scale_x_continuous(name = "Length", breaks = 1:7) +
  scale_y_continuous(name = "Width", breaks = seq(0,3,0.5)) +
  theme(axis.text.x = element_text(face="bold", color="#993333", size=14, angle=45), 
        axis.text.y = element_text(face="bold", color="#993333", size=14, angle=45))
```

It is also possible hide the `ticks`.

```{r, warning=FALSE, message=FALSE}
ggplot(data=iris,mapping=aes(x=Petal.Length,y=Petal.Width)) +
  geom_point(aes(color=Species,size=Sepal.Width)) +
  scale_x_continuous(name = "Length", breaks = 1:7) +
  scale_y_continuous(name = "Width", breaks = seq(0,3,0.5)) +
  theme(axis.text.x = element_text(face="bold", color="#993333", size=14, angle=45), 
        axis.text.y = element_text(face="bold", color="#993333", size=14, angle=45), 
        axis.ticks = element_blank())
```

There are many things one can use to style the axis and/or axis labels. Just use `?theme()` to look for all the different one can use to stylize the plots.

## Different themes

Let's consider the plot below and save it as an object `P` for the sake of simplicity.

```{r, plot.fold = FALSE, warning=FALSE, message=FALSE}
P <- gc_long %>% 
  group_by(Time, Replicate) %>% 
  summarise(mean=mean(log10(count +1)),se=se(log10(count +1))) %>%
  ggplot(aes(x= Time, y= mean, fill = Replicate)) + 
  geom_col(position = position_dodge2()) +
  geom_errorbar(aes(ymin=mean-se, ymax=mean+se), position = position_dodge2(.9, padding = .6)) +
  theme(axis.ticks = element_blank())
P
```

### Lighter themes

`theme_light()`, `theme_minimal()`, `theme_classic()` and `theme_bw()` are a couple of themes that are used very often in publications.

```{r}
P + theme_bw(base_size = 16)
```

```{r}
P + theme_minimal(base_size = 16)
```

### ggthemes package

Let's look into some of the fancier themes that comes in this package

```{r, plot.fold = FALSE, warning=FALSE, message=FALSE}
Q <- ggplot(data = gc_long, mapping = aes(x = Sample_Name, y = log10(count + 1), fill = Time)) + geom_boxplot()
Q  
```

Using the `theme_tufte()`

```{r}
library(ggthemes)
Q + theme_tufte()
```


```{r}
Q + theme_economist() + 
  scale_color_economist()
```

```{r}
Q + theme_stata() +
  scale_color_stata()
```

# Exercise

<i class="fas fa-clipboard-list"></i> Task &nbsp; Try to replicate the plot below if you have enough time.

```{r, plot.fold=FALSE, echo=FALSE, warning=FALSE, message=FALSE}
gc_long %>% 
  group_by(Time, Replicate) %>% 
  summarise(mean=mean(log10(count +1)),se=se(log10(count +1))) %>%
  ggplot(aes(x=Time, y=mean, color = Replicate)) +
  facet_wrap(~Replicate) +
  geom_line(aes(group=1), stat= "identity", size = 2) +
  scale_x_discrete(limits= c("t0", "t2", "t24")) +
  scale_y_continuous(limits = c(0.4,0.8), breaks = seq(0.4,0.8,0.05)) +
  guides(color="none") +
  ylab(label = "mean(log10(count + 1))") +
  theme_light() +
  theme(axis.text = element_text(face="bold", size=12),
        axis.title = element_text(face="bold", color = "#C84DF9", size=14),
        axis.ticks = element_blank())
```

<i class="fas fa-lightbulb"></i> Tip: 1 &nbsp; `geom_line()` is a bit tricky when you use it together with `groups`. It tries to draw lines within the group. In this case, if you would like to draw lines between the groups (like in the above plot, between `t0` through `t2` to `t24`), you initate the ggplot with aesthetics for the line and add `geom_line(aes(group=1))` this way. 

<i class="fas fa-lightbulb"></i> Tip: 2 &nbsp; This figure has `theme_light()`



# Session info

```{r, fold.output=FALSE, fold.plot=FALSE}
sessionInfo()
```

***

