---
title: "ggplot: Part II"
subtitle: "Workshop on Plotting in R"
author: "`r paste0('<b>Lokesh Mano</b> • ',format(Sys.time(), '%d-%b-%Y'))`"
---

```{r, include=FALSE}
hooks = knitr::knit_hooks$get()
hook_foldable = function(type) {
  force(type)
  function(x, options) {
    res = hooks[[type]](x, options)
    
    if (isFALSE(options[[paste0("fold.", type)]])) return(res)
    
    paste0(
      "<details><summary>", type, "</summary>\n\n",
      res,
      "\n\n</details>"
    )
  }
}
knitr::knit_hooks$set(
  output = hook_foldable("output"),
  plot = hook_foldable("plot")
)
```

```{r,child="assets/header-lab.Rmd"}
```


```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(dplyr)
library(ggplot2)
library(reshape2)
data("iris")
gc <- read.table("data/counts_raw.txt", header = T, row.names = 1, sep = "\t")
md <- read.table("data/metadata.csv", header = T, sep = ";")
rownames(md) <- md$Sample_ID

library(tidyverse)
gc_long <- gc %>%
  rownames_to_column(var = "Gene") %>%
  gather(Sample_ID, count, -Gene) %>% 
  full_join(md, by = "Sample_ID") %>% 
  select(Sample_ID, everything()) %>% 
  select(-c(Gene,count), c(Gene,count)) 
gc_long$Sample_Name <- factor(gc_long$Sample_Name, levels = c("t0_A","t0_B","t0_C","t2_A","t2_B","t2_C","t6_A","t6_B","t6_C","t24_A","t24_B","t24_C"))
gc_long$Time <- factor(gc_long$Time, levels = c("t0","t2","t6","t24"))
gc_long$Replicate <- factor(gc_long$Replicate, levels = c("A","B","C"))
```

# Faceting

## With wrap 

We can create subplots using the faceting functionality.

```{r}
ggplot(data=iris,mapping=aes(x=Petal.Length,y=Petal.Width))+
  geom_point(aes(color=Sepal.Width))+
  geom_smooth(method="lm") + 
  facet_wrap(~Species)
```

If we try the same with the gene counts data faceted by time.

```{r}
ggplot(data = gc_long) +
  geom_boxplot(mapping = aes(x = Sample_Name, y = log10(count + 1), color = Time)) +
  facet_wrap(~Time)
```


## With grid

Here in the above plot, you see some empty samples in each facet. In this case, you could use `facet_grid` together with `space` and `scales` options to make it look neat and intuitive. You can use `?facet_grid` and `?facet_wrap` to figure out the exact difference between the two.


```{r}
ggplot(data = gc_long) +
  geom_boxplot(mapping = aes(x = Sample_Name, y = log10(count + 1), color = Time)) +
  facet_grid(~Time , scales = "free", space = "free")
```

You can also make grid with different variables one might have using `vars()` function together with `rows` and `cols` options! 

```{r}
ggplot(data = gc_long) +
  geom_boxplot(mapping = aes(x = Sample_Name, y = log10(count + 1), color = Time)) +
  facet_grid(rows = vars(Time), cols = vars(Replicate), scales = "free", space = "free")
```

# Labeling and annotations

## Labels

Here, we will quickly mention, how one can add labels to the plots. Items on the plot can be labelled using the `geom_text` or `geom_label` geoms.

```{r}
ggplot(data=iris,mapping=aes(x=Petal.Length,y=Petal.Width))+
  geom_point(aes(color=Species))+
  geom_text(aes(label=Species,hjust=0),nudge_x=0.5,size=3)
```

```{r}
ggplot(data=iris,mapping=aes(x=Petal.Length,y=Petal.Width))+
  geom_point(aes(color=Species))+
  geom_label(aes(label=Species,hjust=0),nudge_x=0.5,size=3)
```

The R package ggrepel allows for non-overlapping labels.

```{r}
library(ggrepel)
ggplot(data=iris,mapping=aes(x=Petal.Length,y=Petal.Width))+
  geom_point(aes(color=Species))+
  geom_text_repel(aes(label=Species),size=3)
```

## Annotations

Custom annotations of any geom can be added arbitrarily anywhere on the plot.

```{r}
ggplot(data=iris,mapping=aes(x=Petal.Length,y=Petal.Width))+
  geom_point(aes(color=Species))+
  annotate("text",x=2.5,y=2.1,label="There is a random line here")+
  annotate("segment",x=2,xend=4,y=1.5,yend=2)
```

# Bar charts

Let's now make some bar charts with the data we have. We can start with the simple `iris` data first.

```{r}
ggplot(data=iris,mapping=aes(x=Species,y=Petal.Width))+
  geom_col()
```

<i class="fas fa-exclamation-circle"></i> Note &nbsp; There are two types of bar charts: `geom_bar()` and `geom_col()`. `geom_bar()` makes the height of the bar proportional to the number of cases in each group (or if the `weight` aesthetic is supplied, the sum of the weights). If you want the heights of the bars to represent values in the data, use `geom_col()` instead. `geom_bar()` uses `stat_count()` by default: it counts the number of cases at each x position. `geom_col()` uses `stat_identity()` and it leaves the data as is.

Similarly, we can use the `gene counts` data to make a barplot as well. But first, let's make the data into the right format so as to make the bar plots. This is where knowledge on `tidyverse` would be super useful. 


```{r}
se <- function(x) sqrt(var(x)/length(x))
gc_long %>% 
  group_by(Time) %>% 
  summarise(mean=mean(log10(count +1)),se=se(log10(count +1))) %>%
  head()
```

<i class="fas fa-exclamation-circle"></i> Note &nbsp; There are a couple of things to note here. In the above example, we use the pipe `%>%` symbol that redirects the output of one command as the input to another. Then we group the data by the variable `Time`, followed by summarizing the `count`  with `mean()` and `sd()` functions to get the mean and standard deviation of their respective counts. The `head()` function just prints the first few lines.

Now that we have summarized the data to be bale to plot the bar graph that we want, we can just input the data to ggplot as well using the `%>%` sign. 

```{r}
gc_long %>% 
  group_by(Time) %>% 
  summarise(mean=mean(log10(count +1)),se=se(log10(count +1))) %>%
  ggplot(aes(x=Time, y=mean)) + 
  geom_bar(stat = "identity")
```

<i class="fas fa-exclamation-circle"></i> Note &nbsp; Notice that the `%>%` sign is used in the `tidyverse` based commands and `+` is used for all the `ggplot` based commands.

## Flip coordinates

One can also easily just flip the `x` and `y` axis.

```{r}
gc_long %>% 
  group_by(Time) %>% 
  summarise(mean=mean(log10(count +1)),se=se(log10(count +1))) %>%
  ggplot(aes(x=Time, y=mean)) + 
  geom_col() +
  coord_flip()
```

# Error bars

Now that we have the bar plots, we can also add error bars to them using the `sd` values we calculated in the previous step.

```{r}
gc_long %>% 
  group_by(Time) %>% 
  summarise(mean=mean(log10(count +1)),se=se(log10(count +1))) %>%
  ggplot(aes(x=Time, y=mean, fill = Time)) + 
  geom_col() +
  geom_errorbar(aes(ymax=mean+se,ymin=mean-se),width=0.2)
```


# Stacked bars

Let's now try to make stacked bars. For this let's try to make the data more usable for stacked bars. For this let's use the `group_by` function to make the groups based on both `Time` and `Replicate`.

```{r}
se <- function(x) sqrt(var(x)/length(x))
gc_long %>% 
  group_by(Time, Replicate) %>% 
  summarise(mean=mean(log10(count +1)),se=se(log10(count +1))) %>%
  head()
```

Let's build the stacked bars!

```{r}
gc_long %>% 
  group_by(Time, Replicate) %>% 
  summarise(mean=mean(log10(count +1)),se=se(log10(count +1))) %>%
  ggplot(aes(x=Time, y=mean, fill = Replicate)) + 
  geom_col(position = "stack")
```

One can also have `dodge` bars.

```{r}
gc_long %>% 
  group_by(Time, Replicate) %>% 
  summarise(mean=mean(log10(count +1)),se=se(log10(count +1))) %>%
  ggplot(aes(x=Time, y=mean, fill = Replicate)) + 
  geom_col(position = "dodge")
```

We can try now to plot error bars on them. The errorbars would look weird and complicated if one forgets to add `position = dodge` to the `geom_errorbar()` as well.

```{r, message=FALSE, warning=FALSE}
gc_long %>% 
  group_by(Time, Replicate) %>% 
  summarise(mean=mean(log10(count +1)),se=se(log10(count +1))) %>%
  ggplot(aes(x= Time, y= mean, fill = Replicate)) + 
  geom_col(position = "dodge") +
  geom_errorbar(aes(ymin=mean-se, ymax=mean+se), position = "dodge")
```

<i class="fas fa-exclamation-circle"></i> Note &nbsp; It is important that you keep tract of what kind of aesthetics you give when you initialize `ggplot()` and what you add in the `geoms()` later. 

You can also make these error bars look nicer by playing around with some of the parameters available, like example below:

```{r}
gc_long %>% 
  group_by(Time, Replicate) %>% 
  summarise(mean=mean(log10(count +1)),se=se(log10(count +1))) %>%
  ggplot(aes(x= Time, y= mean, fill = Replicate)) + 
  geom_col(position = position_dodge2()) +
  geom_errorbar(aes(ymin=mean-se, ymax=mean+se), position = position_dodge2(.9, padding = .6))
```

# Exercise

<i class="fas fa-clipboard-list"></i> Task &nbsp; Make the following plots.

<i class="fas fa-lightbulb"></i> Tip &nbsp; It is more of a `tidyverse` exercise than `ggplot`. Because to get these plots, you need get the data in the right format.

<i class="fas fa-clipboard-list"></i> Task &nbsp; Plot 1:

```{r, echo=FALSE, warning=FALSE, message=FALSE, fold.plot=FALSE}

gc_raw <- read.table(file = "data/counts_raw.txt", sep = "\t", header = T)
gc_filt <- read.table(file = "data/counts_filtered.txt", sep = "\t", header = T)
gc_vst <- read.table(file = "data/counts_vst.txt", sep = "\t", header = T)
gc_deseq <- read.table(file = "data/counts_deseq2.txt", sep = "\t", header = T)
md <- read.table("data/metadata.csv", header = T, sep = ";")

gene_counts_all <- 
  gc_raw %>% gather(Sample_ID, Raw, -Gene) %>%
  full_join(gc_filt %>% gather(Sample_ID, Filtered, -Gene), by = c("Gene", "Sample_ID")) %>%
  full_join(gc_vst %>% gather(Sample_ID, VST, -Gene), by = c("Gene", "Sample_ID")) %>%
  full_join(gc_deseq %>% gather(Sample_ID, DESeq2, -Gene), by = c("Gene", "Sample_ID")) %>%
  gather(Method, count, Raw:DESeq2) %>%
  filter(!is.na(count)) %>%
  full_join(md, by = "Sample_ID")

gene_counts_all$Time <- factor(gene_counts_all$Time, levels = c("t0","t2","t6","t24"))
gene_counts_all$Replicate <- factor(gene_counts_all$Replicate, levels = c("A","B","C"))
gene_counts_all$Method <- factor(gene_counts_all$Method, levels = c("Raw","Filtered","DESeq2","VST"))

gene_counts_all %>% 
  group_by(Time, Replicate, Method) %>% 
  summarise(mean=mean(log10(count +1)),se=se(log10(count +1))) %>%
  ggplot(aes(x= Time, y= mean, fill = Replicate)) + 
  geom_bar(position = position_dodge2(), stat = "identity") +
  geom_errorbar(aes(ymin=mean-se, ymax=mean+se), position = position_dodge2(.9, padding = .6)) +
  facet_wrap(~Method, scales = "free")
```

<i class="fas fa-clipboard-list"></i> Task &nbsp; Plot 2:

```{r, echo=FALSE, warning=FALSE, message=FALSE, fold.plot=FALSE, fig.width = 10, fig.height=8}
gene_counts_all %>% 
  group_by(Time, Replicate, Method) %>% 
  ggplot() +
  geom_boxplot(mapping = aes(x = Sample_Name, y = log10(count + 1), fill = Time)) +
  facet_wrap(~Method*Replicate, ncol = 3, scales = "free")
```

# Session info

```{r, fold.output=FALSE, fold.plot=FALSE}
sessionInfo()
```

***

