---
title: "ggplot Part IV"
subtitle: "Workshop on Plotting in R"
author: "`r paste0('<b>Lokesh Mano</b> • ',format(Sys.time(), '%d-%b-%Y'))`"
institute: NBIS, SciLifeLab
keywords: bioinformatics, course, workshop, scilifelab, nbis
output:
  xaringan::moon_reader:
    encoding: 'UTF-8'
    self_contained: false
    chakra: 'assets/remark-latest.min.js'
    css: 'assets/slide.css'
    lib_dir: libs
    nature:
      ratio: '4:3'
      highlightLanguage: r
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      slideNumberFormat: "%current%/%total%"
    include: NULL
---
exclude: true
count: false

```{r,echo=FALSE,child="assets/header-slide.Rmd"}
```

<!-- ------------ Only edit title, subtitle & author above this ------------ -->


```{r,include=FALSE}
# load the packages you need
library(dplyr)
library(tidyr)
library(tidyverse)
#library(stringr)
library(ggplot2)
#library(plotly)
#library(pheatmap)
#library(DESeq2)
#library(edgeR)
library(kableExtra)
library(gridExtra)
```

---
name: save

## Saving plots

```{r,eval=FALSE}
p <- ggplot(iris,aes(Petal.Length,Sepal.Length,color=Species))+
  geom_point()
```

* `ggplot2` plots can be saved just like base plots  

```{r,eval=FALSE}
png("plot.png",height=5,width=7,units="cm",res=200)
print(p)
dev.off()
```

* `ggplot2` package offers a convenient function  

```{r,eval=FALSE}
ggsave("plot.png",p,height=5,width=7,units="cm",dpi=200,type="cairo")
```

* Use `type="cairo"` for nicer anti-aliasing  
* Note that default units in `png` is pixels while in `ggsave` it's inches  

---
name: extension
class: spaced

## Extensions

* [**gridExtra**](https://cran.r-project.org/web/packages/gridExtra/index.html): Extends grid graphics functionality
* [**ggpubr**](http://www.sthda.com/english/rpkgs/ggpubr/): Useful functions to prepare plots for publication  
* [**cowplot**](https://cran.r-project.org/web/packages/cowplot/vignettes/introduction.html): Combining plots  
* [**ggthemes**](https://cran.r-project.org/web/packages/ggthemes/vignettes/ggthemes.html): Set of extra themes  
* [**ggthemr**](https://github.com/cttobin/ggthemr): More themes  
* [**ggsci**](https://cran.r-project.org/web/packages/ggsci/vignettes/ggsci.html): Color palettes for scales  
* [**ggrepel**](https://cran.r-project.org/web/packages/ggrepel/vignettes/ggrepel.html): Advanced text labels including overlap control  
* [**ggmap**](https://github.com/dkahle/ggmap): Dedicated to mapping  
* [**ggraph**](https://github.com/thomasp85/ggraph): Network graphs  
* [**ggiraph**](http://davidgohel.github.io/ggiraph/): Converting ggplot2 to interactive graphics  


---
name: help
class: spaced

## Help

* [**ggplot2 official reference**](http://ggplot2.tidyverse.org/reference/)  
* [**The R cookbook**](http://www.cookbook-r.com/)  
* [**StackOverflow**](https://stackoverflow.com/)  
* [**RStudio Cheatsheet**](https://www.rstudio.com/resources/cheatsheets/)  
* [**r-statistics Cheatsheet**](http://r-statistics.co/ggplot2-cheatsheet.html)  
* [**ggplot2 GUI**](https://site.shinyserver.dck.gmw.rug.nl/ggplotgui/)  
* Numerous personal blogs, r-bloggers.com etc.  

<!-- --------------------- Do not edit this and below --------------------- -->

---
name: end-slide
class: end-slide, middle
count: false


# Thank you. Questions?

```{r,echo=FALSE,child="assets/footer-slide.Rmd"}
```

```{r,include=FALSE,eval=FALSE}
# manually run this to render this document to HTML
rmarkdown::render("slide_sample.Rmd")
# manually run this to convert HTML to PDF
#pagedown::chrome_print("slide_sample.html",output="slide_sample.pdf")
```
