---
title: "R Shiny - Deployments"
subtitle: "Workshop on Plotting in R"
author: "`r paste0('<b>Lokesh Mano</b> • ',format(Sys.time(), '%d-%b-%Y'))`"
institute: NBIS, SciLifeLab
keywords: bioinformatics, course, workshop, scilifelab, nbis
output:
  xaringan::moon_reader:
    encoding: 'UTF-8'
    self_contained: false
    chakra: 'assets/remark-latest.min.js'
    css: 'assets/slide.css'
    lib_dir: libs
    nature:
      ratio: '4:3'
      highlightLanguage: r
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      slideNumberFormat: "%current%/%total%"
    include: NULL
---
exclude: true
count: false

```{r,echo=FALSE,child="assets/header-slide.Rmd"}
```

<!-- ------------ Only edit title, subtitle & author above this ------------ -->

```{r,include=FALSE}
# load the packages you need
library(dplyr)
library(tidyr)
library(tidyverse)
#library(stringr)
library(ggplot2)
#library(plotly)
#library(pheatmap)
#library(DESeq2)
#library(edgeR)
library(kableExtra)
```

---
name: deploy

## Deployment

* R scripts
  * GitHub
  * R Package
  * Docker container

* Webpage
  * Shinyapps.io
  * Shiny Server (Free)
  * Shiny Server Pro

--

* Automatically deploying to shinyapps.io

```
library(rsconnect)
rsconnect::setAccountInfo(name="username", token="HDFGT46YF7TDT6474G47", secret="hdgTDF5FStgdkpJ")
deployApp(appName="name")
```

--

* Shiny apps user guide https://docs.rstudio.com/shinyapps.io/index.html  
* Shiny server guide https://docs.rstudio.com/shiny-server/

---
name: interactive-docs

## Interactive documents

* Shiny can run in RMarkdown documents. Set YAML `runtime: shiny`.  

```
---
title: Interactive document
output: html_document
runtime: shiny
---
```

* Shiny widgets can be included directly  

```{r,echo=FALSE,comment=""}
cat(paste0("```{r}\n","selectInput('n_breaks',label='Number of bins:',choices=c(10,20,35,50),selected=20)\n","```\n"))
```

* Whole shiny apps can be included directly  

```{r,echo=FALSE,comment=""}
cat(paste0("```{r}\n","shinyApp(\n","  ui=fluidPage(),\n","  server=function(input,output) {}\n",")\n","```\n"))
```

* Hosted shiny apps can be embedded using `<iframe>`  

```
<iframe src="https://user.shinyapps.io/app"></iframe>
```

.small[_Demo: shiny-rmarkdown.R_]

---
name: extensions

## Extensions

* [Naxstats](https://github.com/nanxstats/awesome-shiny-extensions): Repo collection all shiny extensions
* [shinythemes](https://github.com/rstudio/shinythemes): Bootswatch themes for shiny
* shinyurl: using URLs to recreate state of an app
* shinypod: Reusable modules
* [shinyjs](https://deanattali.com/shinyjs/): custom javascript functionality
* [shinyWidgets](https://github.com/dreamRs/shinyWidgets): Bootstrap 3 custom widgets
* [shinyBS](https://github.com/ebailey78/shinyBS): Bootstrap 3 widgets

---
name: help
class: spaced

## Help

* [**RStudio Shiny home**](https://shiny.rstudio.com/) 
* [**RStudio Shiny tutorial**](http://shiny.rstudio.com/tutorial/)
* [**RStudio Shiny articles**](https://shiny.rstudio.com/articles/)

### Other topics

* [Applications of action buttons](http://shiny.rstudio.com/articles/action-buttons.html)  
* [Progress indicators](http://shiny.rstudio.com/articles/progress.html)  
* [Shiny modules](https://shiny.rstudio.com/articles/modules.html)
...
---

name: end_slide
class: end-slide, middle
count: false

# Thank you. Questions?

Slide courtesy: Roy Francis (NBIS, RaukR2021)

```{r,echo=FALSE,child="assets/footer-slide.Rmd"}
```

```{r,include=FALSE,eval=FALSE}
# manually run this to render this document to HTML
rmarkdown::render("slide_sample.Rmd")
# manually run this to convert HTML to PDF
#pagedown::chrome_print("slide_sample.html",output="slide_sample.pdf")
```
