---
title: "R Shiny - Deployments"
author: Lokesh Mano
institute: NBIS, SciLifeLab
keywords: bioinformatics, course, workshop, scilifelab, nbis
format: 
  revealjs: 
    slide-number: true
---

## Setup {visibility="hidden"}

```{r,include=FALSE}
# load the packages you need
library(dplyr)
library(tidyr)
library(tidyverse)
#library(stringr)
library(ggplot2)
#library(plotly)
#library(pheatmap)
#library(DESeq2)
#library(edgeR)
library(kableExtra)
```

## Deployment

:::: {.fragment}

* R scripts
  * GitHub
  * R Package
  * Docker container

::::

:::: {.fragment}

* Webpage
  * Shinyapps.io
  * Shiny Server (Free)
  * Posit Connect

:::: 

:::: {.fragment}

* Automatically deploying to shinyapps.io

::::

## Shinyapps.io

```{r, eval=FALSE}
library(rsconnect)
rsconnect::setAccountInfo(name="username", token="HDFGT46YF7TDT6474G47", secret="hdgTDF5FStgdkpJ")
deployApp(appName="name")
```

:::: {.fragment}

* [Shiny apps user guide](https://docs.posit.co/shinyapps.io/guide/index.html)
* [Shiny server guide](https://posit.co/products/open-source/shiny-server/)

::::

## Interactive documents

* Shiny can run in RMarkdown documents. Set YAML `runtime: shiny`.  

:::: {.fragment}

```
---

title: Interactive document
output: html_document
runtime: shiny
---

```

:::: 

:::: {.fragment}

* Shiny widgets can be included directly  

```{r,echo=FALSE,comment=""}
cat(paste0("```{r}\n","selectInput('n_breaks',label='Number of bins:',choices=c(10,20,35,50),selected=20)\n","```\n"))
```

::::

## Interactive documents

* Whole shiny apps can be included directly  

```{r,echo=FALSE,comment=""}
cat(paste0("```{r}\n","shinyApp(\n","  ui=fluidPage(),\n","  server=function(input,output) {}\n",")\n","```\n"))
```

:::: {.fragment}

* Hosted shiny apps can be embedded using `<iframe>`  

```
<iframe src="https://user.shinyapps.io/app"></iframe>
```

::::

## Extensions

* [Naxstats](https://github.com/nanxstats/awesome-shiny-extensions): Repo collection all shiny extensions
* [shinythemes](https://github.com/rstudio/shinythemes): Bootswatch themes for shiny
* shinyurl: using URLs to recreate state of an app
* shinypod: Reusable modules
* [shinyjs](https://deanattali.com/shinyjs/): custom javascript functionality
* [shinyWidgets](https://github.com/dreamRs/shinyWidgets): Bootstrap 3 custom widgets
* [shinyBS](https://github.com/ebailey78/shinyBS): Bootstrap 3 widgets

## Help

* [**Posit Shiny home**](https://shiny.posit.co/) 
* [**Posit Shiny tutorial**](https://shiny.posit.co/r/getstarted/shiny-basics/lesson1/)
* [**Posit Shiny articles**](https://shiny.posit.co/r/articles/)
* [**Shiny using python**](https://shiny.posit.co/py/docs/overview.html)

### Other topics

* [Applications of action buttons](https://shiny.posit.co/r/articles/build/action-buttons/)  
* [Progress indicators](https://shiny.posit.co/r/articles/build/progress/)  
* [Shiny modules](https://shiny.posit.co/r/articles/improve/modules/)


## {background-image="../../assets/images/precourse/data_viz_adv_2.jpeg"}

::: {.center-xy}

### Thank you. Questions? {style="text-align: center;"}
Slide inspirations: Roy Francis (NBIS, RaukR2024)

:::

