---
title: "Basics in ggplot"
subtitle: "Workshop on Advanced Data Visualization"
author: "Lokesh Mano"
format: 
    html: 
        font-size: 10
code-annotations: hover
---

```{r,include=FALSE}
library(dplyr)
library(ggplot2)
library(kableExtra)
library(tidyr)
library(tidyverse)
```

We will use R in this course for generating plots from different biological data that are of higher quality and standard for publications. To do this, you will brush-up your memory on some important aspects of R that are important for this course below:

:::{.callout-caution title="Download course files"}
Before starting with the lab sessions of the entire course, you must have downloaded all the necessary files from [here](https://nbisweden.github.io/workshop-data-visualization-r/2505/topics/lab_download.html) and make sure that the directory tree looks similar to the one displayed in that page

In that case, you can proceed with the exercise now and remember to have fun 😊
:::

:::{.callout-note}
Throughout all the materials in the course the files are read in R with location like in the example below:
```{r, eval=FALSE}
df <- read.table("../../data/counts_raw.txt")
```

This is due to how the GitHub repo for this course is designed to be able to render these webpages. 
For you, depending on where your current working location on the R terminal is, you might have to change the relative location of the `data` directory accordingly. For example, if `data` directory is present when you run `dir()` in the R terminal, you can just read-in the files to R like following:

```{r, eval=FALSE}
df <- read.table("data/counts_raw.txt")
```

Otherwise, you can set the current working directory to the place where you will have all the course-related materials in your computer. You only have to do this once per R session. An example is shown below how this looks in my computer:

```{r, eval=FALSE}
setwd('/home/lokesh/workshop_on_advanced_data_visualization/')
df <- read.table("data/counts_raw.txt")
```

:::

# Data formats 

In terms of the R (and other similar programming languages), the data can be viewed or stored in two main formats! They are called `wide` and `long` formats. Below you will see what exactly they stand for and why is it important for plotting in ggplot.

## Wide format

A quick preview: 

**Counts Table**

```{r, echo=FALSE}
gc <- read.table("../../data/counts_raw.txt", header = T, row.names = 1, sep = "\t")
kable(gc[c(1:6),c(1:4)]) %>%
  kable_styling(bootstrap_options = "striped", full_width = F)
```

And we usually have our metadata related to the samples in another table like below:

**Metadata Table**

```{r, echo=FALSE}
md <- read.table("../../data/metadata.csv", header = T, sep = ";")
kable(md[c(1:4),]) %>%
  kable_styling(bootstrap_options = "striped", full_width = F)
```

* Wide format data is called “wide” because it typically has a lot of columns which stretch widely across the page or your computer screen.
* Most of us are familiar with looking at wide format data
  + It is convenient and we are more used to looking at data this way in our Excel sheets. 
  + It often lets you see more of the data, at one time, on your screen

## Long format

Below is glimpse how the long format of the same data look like:

```{r echo=FALSE}
samples <- colnames(gc[,c(1:4)])
gc[c(1:6),c(1:4)] %>%
  rownames_to_column(var = "Gene") %>%
  gather(Samples, count, -Gene) %>% 
  head(10) %>%
  kable() %>%
  kable_styling(bootstrap_options = "striped", full_width = F)
```

Or to be even more complete and precise:

```{r echo=FALSE}
samples <- colnames(gc[,c(1:4)])
gc[c(1:6),c(1:4)] %>%
  rownames_to_column(var = "Gene") %>%
  gather(Sample_ID, count, -Gene) %>% 
  full_join(md[c(1:4),], by = "Sample_ID") %>% 
  select(Sample_ID, everything()) %>% 
  select(-c(Gene,count), c(Gene,count)) %>%
  head(10) %>%
  kable() %>%
  kable_styling("striped", full_width = F)
```

* Long format data is typically less familiar to most humans
  + It seems awfully hard to get a good look at all (or most) of it
  + It seems like it would require more storage on your hard disk
  + It seems like it would be harder to enter data in a long format

## Which is better?

* Well, there are some contexts where putting things in wide format is computationally efficient because you can treat data in a matrix format and to efficient matrix calculations on it.
* However, adding data to wide format data sets is very hard:
  1. It is very difficult to conceive of analytic schemes that apply generally across all wide-format data sets.
  2. Many tools in R want data in long format **like ggplot**
  3. The long format for data corresponds to the relational model for storing data, which is the model used in most modern data bases like the SQL family of data base systems.
  
* A more technical treatment of wide versus long data requires some terminology:
  - <span style="color:blue">Identifier variables</span> are often categorical things that cross-classify observations into categories.
  - <span style="color:red">Measured variables</span> are the names given to properties or characteristics that you can go out and measure.
  - <span style="color:orange">Values</span> are the values that you measure are record for any particular measured variable.

* In any particular data set, what you might want to call an <span style="color:blue">Identifier variables</span> versus a <span style="color:red">Measured variables</span> can not always be entirely clear.
  - Other people might choose to define things differently.
* However, to my mind, it is less important to be able to precisely recognize these three entities in every possible situation (where there might be some fuzziness about which is which)
* And it is more important to understand how <span style="color:blue">Identifier variables</span>, <span style="color:red">Measured variables</span>, and <span style="color:orange">Values</span> interact and behave when we are transforming data between wide and long formats.


# Conversion between formats

As for the biological data analysis, to be able to use tools such as **ggplot**, in simple terms we should learn to convert our data:

As per our previous examples: We should learn to convert 

**From this format**

```{r, echo=FALSE}
gc <- read.table("../../data/counts_raw.txt", header = T, row.names = 1, sep = "\t")
kable(gc[c(1:6),c(1:4)]) %>%
  kable_styling(bootstrap_options = "striped", full_width = F) %>%
  row_spec(1:6, color = "orange") %>%
  column_spec(1, color = "red") %>%
  row_spec(0, bold = T, color = "blue")
```

**To this format**

```{r echo=FALSE}
samples <- colnames(gc[,c(1:4)])
gc[c(1:6),c(1:4)] %>%
  rownames_to_column(var = "Gene") %>%
  gather(Sample_ID, count, -Gene) %>% 
  full_join(md[c(1:4),], by = "Sample_ID") %>% 
  select(Sample_ID, everything()) %>% 
  select(-c(Gene,count), c(Gene,count)) %>%
  head(10) %>%
  kable() %>%
  kable_styling("striped", full_width = F) %>%
  column_spec(1:5, color = "blue") %>%
  column_spec(6, color = "red")%>%
  column_spec(7, color = "orange")
```

Here we will only cover the conversion from `wide` to `long`, as this is more relevant to us. For the other way around, one can look into `spread()` from the `tidyr` package.

## Using reshape2

By using the `melt()` function from the **reshape2** package we can convert the wide-formatted data into long-formatted data! Here, to combine the metadata table to the gene counts table, we will also use the `merge()` function like we did before!

```{r}
library(reshape2)
gc <- read.table("../../data/counts_raw.txt", header = T, row.names = 1, sep = "\t")
md <- read.table("../../data/metadata.csv", header = T, sep = ";")
rownames(md) <- md$Sample_ID

#merging gene counts table with metadata
merged_data_wide <- merge(md, t(gc), by = 0)
#removing redundant columns
merged_data_wide$Row.names <- NULL

merged_data_long <- melt(merged_data_wide, id.vars = c("Sample_ID","Sample_Name","Time","Replicate","Cell"), variable.name = "Gene", value.name = "count")  
head(merged_data_long)
```

## Using tidyr

If you are more familiar with using `tidyverse` or `tidyr` packages, you can combine tables by `join()` and then use `gather()` to make long formatted data in the same command. This is a powerful and more cleaner way of dealing with data in R.

```{r, warning=FALSE, message=FALSE}
library(tidyverse)
gc_long <- gc %>%
  rownames_to_column(var = "Gene") %>%
  gather(Sample_ID, count, -Gene) %>% 
  full_join(md, by = "Sample_ID") %>% 
  select(Sample_ID, everything()) %>% 
  select(-c(Gene,count), c(Gene,count)) 
gc_long %>%
  head(10)
```

# Exercise I

:::{.callout-important title="Task 1.1"}
Here in this exercise we have used the `counts_raw.txt` file, you can try to make similar R objects for each of the other three counts (`counts_filtered.txt`, `counts_vst.txt` and `counts_deseq2.txt`) in `long format`. So for example you would have `gc_filt`, `gc_vst` and `gc_deseq2` R objects in the end.
:::

:::{.callout-tip}
Remember to take a look at how these files are formatted, before you import!
:::

# Base vs grid graphics

## Base

R is an excellent tool for creating graphs and plots. The graphic capabilities and functions provided by the base R installation is called the base R graphics. Numerous packages exist to extend the functionality of base graphics.

We can try out plotting a few of the common plot types. Let's start with a scatterplot. First we create a `data.frame` as this is the most commonly used data object.

```{r}
dfr <- data.frame(a=sample(1:100,10),b=sample(1:100,10))
```

Now we have a dataframe with two continuous variables that can be plotted against each other.


```{r}
plot(dfr$a,dfr$b)
```


This is probably the simplest and most basic plots. We can modify the x and y axis labels.

```{r}
plot(dfr$a,dfr$b,xlab="Variable a",ylab="Variable b")
```

We can change the point to a line.

```{r}
plot(dfr$a,dfr$b,xlab="Variable a",ylab="Variable b",type="b")
```

Let's add a categorical column to our dataframe.

```{r}
dfr$cat <- rep(c("C1","C2"),each=5)
```

And then colour the points by category.

```{r}
# subset data
dfr_c1 <- subset(dfr,dfr$cat == "C1")
dfr_c2 <- subset(dfr,dfr$cat == "C2")

plot(dfr_c1$a,dfr_c1$b,xlab="Variable a",ylab="Variable b",col="red",pch=1)
points(dfr_c2$a,dfr_c2$b,col="blue",pch=2)

legend(x="topright",legend=c("C1","C2"),
       col=c("red","blue"),pch=c(1,2))
```

Let's create a barplot.

```{r}
ldr <- data.frame(a=letters[1:10],b=sample(1:50,10))
barplot(ldr$b,names.arg=ldr$a)
```

## Grid

Grid graphics have a completely different underlying framework compared to base graphics. Generally, base graphics and grid graphics cannot be plotted together. The most popular grid-graphics based plotting library is **ggplot2**.

Let's create the same plot as before using **ggplot2**. Make sure you have the package installed.

```{r}
library(ggplot2)

ggplot(dfr)+
  geom_point(mapping = aes(x=a,y=b,colour=cat))+
  labs(x="Variable a",y="Variable b")
```

It is generally easier and more consistent to create plots using the ggplot2 package compared to the base graphics.

Let's create a barplot as well.

```{r}
ggplot(ldr,aes(x=a,y=b))+
  geom_col()
```


## Saving images

Let's take a look at saving plots. 

::: {.callout-note}
This part is just to give you a quick look into how you can save images from R terminal quickly. The different format of images will be explained in a lecture later.
:::

### Base graphics

The general idea for saving plots is open a graphics device, create the plot and then close the device. We will use **png** here. Check out `?png` for the arguments and other devices.

```{r,eval=FALSE}
dfr <- data.frame(a=sample(1:100,10),b=sample(1:100,10))

png(filename="plot-base.png")
plot(dfr$a,dfr$b)
dev.off()
```

### ggplot2

The same idea can be applied to ggplot2, but in a slightly different way. First save the file to a variable, and then export the plot.

```{r,eval=FALSE}
p <- ggplot(dfr,aes(a,b)) + geom_point()

png(filename="plot-ggplot-1.png")
print(p)
dev.off()
```

::: {.callout-tip}
**ggplot2** also has another easier helper function to export images.

```{r,eval=FALSE}
ggsave(filename="plot-ggplot-2.png",plot=p)
```
:::

# ggplot basics

Make sure the library is loaded in your environment.

```{r, eval=FALSE}
library(ggplot2)
```

## Geoms

In the previous section we saw very quickly how to use `ggplot`. Let's take a look at it again a bit more carefully. For this let's first look into a simple data that is available in R. We use the `iris` data for this to start with.

This dataset has four continuous variables and one categorical variable. It is important to remember about the data type when plotting graphs

```{r, fold.output=FALSE}
data("iris")
head(iris)
```

When we initiate the ggplot object using the data, it just creates a blank plot!

```{r}
ggplot(iris) 
```

Now we can specify what we want on the x and y axes using aesthetic mapping. And we specify the geometric using `geoms`. 

::: {.callout-note} 
that the variable names do not have double quotes `""` like in base plots.
:::

```{r}
ggplot(data=iris)+
  geom_point(mapping=aes(x=Petal.Length,y=Petal.Width))
```

### Multiple geoms

Further geoms can be added. For example let’s add a regression line. When multiple geoms with the same aesthetics are used, they can be specified as a common mapping. 

::: {.callout-note} 
that the order in which geoms are plotted depends on the order in which the geoms are supplied in the code. In the code below, the points are plotted first and then the regression line.
:::

```{r}
ggplot(data=iris,mapping=aes(x=Petal.Length,y=Petal.Width))+
  geom_point()+
  geom_smooth(method="lm")
```

There are many other `geoms` and you can find most of them here in this [cheatsheet](../../assets/data-visualization-1.pdf)

### Gene counts data

Let's also try to use `ggplot` for a "more common" gene counts dataset. Let's use the `merged_data_long` or the `gc_long` object we created in the earlier session.

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(reshape2)
gc <- read.table("../../data/counts_raw.txt", header = T, row.names = 1, sep = "\t")
md <- read.table("../../data/metadata.csv", header = T, sep = ";")
rownames(md) <- md$Sample_ID

#merging gene counts table with metadata
merged_data_wide <- merge(md, t(gc), by = 0)
#removing redundant columns
merged_data_wide$Row.names <- NULL

library(tidyverse)
gc_long <- gc %>%
  rownames_to_column(var = "Gene") %>%
  gather(Sample_ID, count, -Gene) %>% 
  full_join(md, by = "Sample_ID") %>% 
  select(Sample_ID, everything()) %>% 
  select(-c(Gene,count), c(Gene,count)) 
```

```{r}
ggplot(data = gc_long) +
  geom_boxplot(mapping = aes(x = Sample_Name, y = log10(count +1)))
```

::: {.callout-note} 
You can notice that the ggplot sorts the `factors` or `variables` alpha-numerically, like in the case above with `Sample_Name`. 
:::

::: {.callout-tip}  
There is a trick that you can use to give the order of variables manually. The example is shown below:

```{r}
gc_long$Sample_Name <- factor(gc_long$Sample_Name, levels = c("t0_A","t0_B","t0_C","t2_A","t2_B","t2_C","t6_A","t6_B","t6_C","t24_A","t24_B","t24_C"))
ggplot(data = gc_long) +
  geom_boxplot(mapping = aes(x = Sample_Name, y = log10(count + 1)))
```
:::

## Colors

### Iris data

First, if we look at the `iris` data, we can use the categorical column `Species` to color the points. The color aesthetic is used by geom_point and geom_smooth. Three different regression lines are now drawn. Notice that a legend is automatically created

```{r}
ggplot(data=iris,mapping=aes(x=Petal.Length,y=Petal.Width,color=Species))+
  geom_point()+
  geom_smooth(method="lm")

```

If we wanted to keep a common regression line while keeping the colors for the points, we could specify color aesthetic only for `geom_point`.

```{r}
ggplot(data=iris,mapping=aes(x=Petal.Length,y=Petal.Width))+
  geom_point(aes(color=Species))+
  geom_smooth(method="lm")

```

### GC data

Similarly, we can do the same with the gene counts data. 

```{r}
ggplot(data = gc_long) +
  geom_boxplot(mapping = aes(x = Sample_Name, y = log10(count + 1), color = Time))
```

:::{.callout-tip}
We can also use the `fill` aesthetic to give it a better look.
:::

```{r}
ggplot(data = gc_long) +
  geom_boxplot(mapping = aes(x = Sample_Name, y = log10(count + 1), fill = Time))
```

### Discrete colors

We can change the default colors by specifying new values inside a scale.

```{r}
ggplot(data=iris,mapping=aes(x=Petal.Length,y=Petal.Width))+
  geom_point(aes(color=Species))+
  geom_smooth(method="lm")+
  scale_color_manual(values=c("red","blue","green"))
```

:::{.callout-tip}
To specify manual colors, you could specify by their `names` or their `hexadecimal codes`. For example, you can choose the colors based on `names` from an online source like in this [cheatsheet](http://www.stat.columbia.edu/~tzheng/files/Rcolor.pdf) or you can use the `hexadecimal code` and choose it from a source like [here](https://htmlcolorcodes.com/). I personally prefer the `hexa` based options for manual colors.
:::

### Continuous colors

We can also map the colors to a continuous variable. This creates a color bar legend item.

```{r}
ggplot(data=iris,mapping=aes(x=Petal.Length,y=Petal.Width))+
  geom_point(aes(color=Sepal.Width))+
  geom_smooth(method="lm")
```

:::{.callout-tip} 
Here, you can also choose different `palettes` for choosing the right continuous palette. There are some common packages of palettes that are used very often. [RColorBrewer](https://www.r-graph-gallery.com/38-rcolorbrewers-palettes.html) and [wesanderson](https://github.com/karthik/wesanderson), if you are a fan of his choice of colors. 😉
:::

```{r}
library(wesanderson)
ggplot(data=iris,mapping=aes(x=Petal.Length,y=Petal.Width))+
  geom_point(aes(color=Sepal.Width))+
  geom_smooth(method="lm") +
  scale_color_gradientn(colours = wes_palette("Moonrise3"))
```

:::{.callout-tip}
You can also use simple R base color palettes like `rainbow()` or `terrain.colors()`. Use `?` and look at these functions to see, how to use them.
:::

## Aesthetics

### Aesthetic parameter

We can change the size of all points by a fixed amount by specifying size outside the aesthetic parameter.

```{r}
ggplot(data=iris,mapping=aes(x=Petal.Length,y=Petal.Width))+
  geom_point(aes(color=Species),size=3)+
  geom_smooth(method="lm")
```

### Aesthetic mapping

We can map another variable as size of the points. This is done by specifying size inside the aesthetic mapping. Now the size of the points denote `Sepal.Width`. A new legend group is created to show this new aesthetic.

```{r}
ggplot(data=iris,mapping=aes(x=Petal.Length,y=Petal.Width))+
  geom_point(aes(color=Species,size=Sepal.Width))+
  geom_smooth(method="lm")
```

# Histogram

Here, as a quick example, we will try to make use of the different combinations of `geoms`, `aes` and `color` in simple plots.

Let's take a quick look at some of widely used functions like histograms and density plots in `ggplot`. Intuitively, these can be drawn with `geom_histogram()` and `geom_density()`. Using `bins` and `binwidth` in `geom_histogram()`, one can customize the histogram.

```{r}
ggplot(data=iris,mapping=aes(x=Sepal.Length))+
  geom_histogram()
```

## Density

Let's look at the sample plot in density.

```{r}
ggplot(data=iris,mapping=aes(x=Sepal.Length))+
  geom_density()
```

The above plot is not very informative, let's see how the different species contribute:

```{r}
ggplot(data=iris,mapping=aes(x=Sepal.Length))+
  geom_density(aes(fill = Species), alpha = 0.8)
```

:::{.callout-note}
The `alpha` option inside `geom_density` controls the transparency of the plot.
:::

# Exercise II

:::{.callout-important title="Task 2.1"}
Make `boxplots` similar to the one we did here in this exercise for the other three counts (`counts_filtered.txt`, `counts_vst.txt` and `counts_deseq2.txt`). 
:::

:::{.callout-tip}
You can save the plots themselves as R objects. You will get the plot by just calling those objects. You can then add layers to those objects. Examples are shown below:
:::

```{r}
plot_obj_1 <- ggplot(data=iris,mapping=aes(x=Petal.Length,y=Petal.Width))+
  geom_point(aes(color=Sepal.Width))+
  geom_smooth(method="lm") 
plot_obj_1
```

```{r}
plot_obj_2 <- plot_obj_1 +
  scale_color_gradientn(colours = wes_palette("Moonrise3"))
plot_obj_2
```

This way, you can create different plot objects for the different counts, we will use them in the later exercises.

# Faceting

## With wrap 

We can create subplots using the faceting functionality.

```{r}
plot <- ggplot(data=iris,mapping=aes(x=Petal.Length,y=Petal.Width))+
  geom_point(aes(color=Sepal.Width))+
  geom_smooth(method="lm") + 
  facet_wrap(~Species)
plot
```

If we try the same with the gene counts data faceted by time.

```{r}
ggplot(data = gc_long) +
  geom_boxplot(mapping = aes(x = Sample_Name, y = log10(count + 1), color = Time)) +
  facet_wrap(~Time)
```


## With grid

Here in the above plot, you see some empty samples in each facet. In this case, you could use `facet_grid` together with `space` and `scales` options to make it look neat and intuitive. You can use `?facet_grid` and `?facet_wrap` to figure out the exact difference between the two.


```{r}
ggplot(data = gc_long) +
  geom_boxplot(mapping = aes(x = Sample_Name, y = log10(count + 1), color = Time)) +
  facet_grid(~Time , scales = "free", space = "free")
```

You can also make grid with different variables one might have using `vars()` function together with `rows` and `cols` options! 

```{r}
ggplot(data = gc_long) +
  geom_boxplot(mapping = aes(x = Sample_Name, y = log10(count + 1), color = Time)) +
  facet_grid(rows = vars(Time), cols = vars(Replicate), scales = "free", space = "free")
```

# Labeling and annotations

## Labels

Here, we will quickly mention, how one can add labels to the plots. Items on the plot can be labelled using the `geom_text` or `geom_label` geoms.

```{r}
ggplot(data=iris,mapping=aes(x=Petal.Length,y=Petal.Width))+
  geom_point(aes(color=Species))+
  geom_text(aes(label=Species,hjust=0),nudge_x=0.5,size=3)
```

```{r}
ggplot(data=iris,mapping=aes(x=Petal.Length,y=Petal.Width))+
  geom_point(aes(color=Species))+
  geom_label(aes(label=Species,hjust=0),nudge_x=0.5,size=3)
```

The R package ggrepel allows for non-overlapping labels.

```{r}
library(ggrepel)
ggplot(data=iris,mapping=aes(x=Petal.Length,y=Petal.Width))+
  geom_point(aes(color=Species))+
  geom_text_repel(aes(label=Species),size=3)
```

## Annotations

Custom annotations of any geom can be added arbitrarily anywhere on the plot.

```{r}
ggplot(data=iris,mapping=aes(x=Petal.Length,y=Petal.Width))+
  geom_point(aes(color=Species))+
  annotate("text",x=2.5,y=2.1,label="There is a random line here")+
  annotate("segment",x=2,xend=4,y=1.5,yend=2)
```

# Bar charts

Let's now make some bar charts with the data we have. We can start with the simple `iris` data first.

```{r}
ggplot(data=iris,mapping=aes(x=Species,y=Petal.Width))+
  geom_col()
```

:::{.callout-note}
There are two types of bar charts: `geom_bar()` and `geom_col()`. `geom_bar()` makes the height of the bar proportional to the number of cases in each group (or if the `weight` aesthetic is supplied, the sum of the weights). If you want the heights of the bars to represent values in the data, use `geom_col()` instead. `geom_bar()` uses `stat_count()` by default: it counts the number of cases at each x position. `geom_col()` uses `stat_identity()` and it leaves the data as is.
:::

Similarly, we can use the `gene counts` data to make a barplot as well. But first, let's make the data into the right format so as to make the bar plots. This is where knowledge on `tidyverse` would be super useful. 


```{r}
se <- function(x) sqrt(var(x)/length(x))
gc_long %>% 
  group_by(Time) %>% 
  summarise(mean=mean(log10(count +1)),se=se(log10(count +1))) %>%
  head()
```

:::{.callout-note}
There are a couple of things to note here. In the above example, we use the pipe `%>%` symbol that redirects the output of one command as the input to another. Then we group the data by the variable `Time`, followed by summarizing the `count`  with `mean()` and `sd()` functions to get the mean and standard deviation of their respective counts. The `head()` function just prints the first few lines.
:::

Now that we have summarized the data to be bale to plot the bar graph that we want, we can just input the data to ggplot as well using the `%>%` sign. 

```{r}
gc_long %>% 
  group_by(Time) %>% 
  summarise(mean=mean(log10(count +1)),se=se(log10(count +1))) %>%
  ggplot(aes(x=Time, y=mean)) + 
  geom_bar(stat = "identity")
```

:::{.callout-note}
Notice that the `%>%` sign is used in the `tidyverse` based commands and `+` is used for all the `ggplot` based commands.
:::

## Flip coordinates

One can also easily just flip the `x` and `y` axis.

```{r}
gc_long %>% 
  group_by(Time) %>% 
  summarise(mean=mean(log10(count +1)),se=se(log10(count +1))) %>%
  ggplot(aes(x=Time, y=mean)) + 
  geom_col() +
  coord_flip()
```

# Error bars

Now that we have the bar plots, we can also add error bars to them using the `sd` values we calculated in the previous step.

```{r}
gc_long %>% 
  group_by(Time) %>% 
  summarise(mean=mean(log10(count +1)),se=se(log10(count +1))) %>%
  ggplot(aes(x=Time, y=mean, fill = Time)) + 
  geom_col() +
  geom_errorbar(aes(ymax=mean+se,ymin=mean-se),width=0.2)
```


# Stacked bars

Let's now try to make stacked bars. For this let's try to make the data more usable for stacked bars. For this let's use the `group_by` function to make the groups based on both `Time` and `Replicate`.

```{r}
se <- function(x) sqrt(var(x)/length(x))
gc_long %>% 
  group_by(Time, Replicate) %>% 
  summarise(mean=mean(log10(count +1)),se=se(log10(count +1))) %>%
  head()
```

Let's build the stacked bars!

```{r}
gc_long %>% 
  group_by(Time, Replicate) %>% 
  summarise(mean=mean(log10(count +1)),se=se(log10(count +1))) %>%
  ggplot(aes(x=Time, y=mean, fill = Replicate)) + 
  geom_col(position = "stack")
```

One can also have `dodge` bars.

```{r}
gc_long %>% 
  group_by(Time, Replicate) %>% 
  summarise(mean=mean(log10(count +1)),se=se(log10(count +1))) %>%
  ggplot(aes(x=Time, y=mean, fill = Replicate)) + 
  geom_col(position = "dodge")
```

We can try now to plot error bars on them. The errorbars would look weird and complicated if one forgets to add `position = dodge` to the `geom_errorbar()` as well.

```{r, message=FALSE, warning=FALSE}
gc_long %>% 
  group_by(Time, Replicate) %>% 
  summarise(mean=mean(log10(count +1)),se=se(log10(count +1))) %>%
  ggplot(aes(x= Time, y= mean, fill = Replicate)) + 
  geom_col(position = "dodge") +
  geom_errorbar(aes(ymin=mean-se, ymax=mean+se), position = "dodge")
```

:::{.callout-note}
It is important that you keep tract of what kind of aesthetics you give when you initialize `ggplot()` and what you add in the `geoms()` later. 
:::

You can also make these error bars look nicer by playing around with some of the parameters available, like example below:

```{r}
gc_long %>% 
  group_by(Time, Replicate) %>% 
  summarise(mean=mean(log10(count +1)),se=se(log10(count +1))) %>%
  ggplot(aes(x= Time, y= mean, fill = Replicate)) + 
  geom_col(position = position_dodge2()) +
  geom_errorbar(aes(ymin=mean-se, ymax=mean+se), position = position_dodge2(.9, padding = .6))
```

# Exercise III

Now let us try to make the following plots as exercise before we go further to some `advanced ggplot` session.

::::{.callout-important title="Task 3.1"}

```{r, echo=FALSE, warning=FALSE, message=FALSE}

gc_raw <- read.table(file = "../../data/counts_raw.txt", sep = "\t", header = T)
gc_filt <- read.table(file = "../../data/counts_filtered.txt", sep = "\t", header = T)
gc_vst <- read.table(file = "../../data/counts_vst.txt", sep = "\t", header = T)
gc_deseq <- read.table(file = "../../data/counts_deseq2.txt", sep = "\t", header = T)
md <- read.table("../../data/metadata.csv", header = T, sep = ";")

gene_counts_all <- 
  gc_raw %>% gather(Sample_ID, Raw, -Gene) %>%
  full_join(gc_filt %>% gather(Sample_ID, Filtered, -Gene), by = c("Gene", "Sample_ID")) %>%
  full_join(gc_vst %>% gather(Sample_ID, VST, -Gene), by = c("Gene", "Sample_ID")) %>%
  full_join(gc_deseq %>% gather(Sample_ID, DESeq2, -Gene), by = c("Gene", "Sample_ID")) %>%
  gather(Method, count, Raw:DESeq2) %>%
  filter(!is.na(count)) %>%
  full_join(md, by = "Sample_ID")

gene_counts_all$Time <- factor(gene_counts_all$Time, levels = c("t0","t2","t6","t24"))
gene_counts_all$Replicate <- factor(gene_counts_all$Replicate, levels = c("A","B","C"))
gene_counts_all$Method <- factor(gene_counts_all$Method, levels = c("Raw","Filtered","DESeq2","VST"))

gene_counts_all %>% 
  group_by(Time, Replicate, Method) %>% 
  summarise(mean=mean(log10(count +1)),se=se(log10(count +1))) %>%
  ggplot(aes(x= Time, y= mean, fill = Replicate)) + 
  geom_bar(position = position_dodge2(), stat = "identity") +
  geom_errorbar(aes(ymin=mean-se, ymax=mean+se), position = position_dodge2(.9, padding = .6)) +
  facet_wrap(~Method, scales = "free")
```

:::{.callout-tip}
It is more of a `tidyverse` exercise than `ggplot`. Because to get these plots, you need get the data in the right format.
:::

::::

::::{.callout-important title="Task 3.2"}

```{r, echo=FALSE, warning=FALSE, message=FALSE, fold.plot=FALSE, fig.width = 10, fig.height=8}
gene_counts_all %>% 
  group_by(Time, Replicate, Method) %>% 
  ggplot() +
  geom_boxplot(mapping = aes(x = Sample_Name, y = log10(count + 1), fill = Time)) +
  facet_wrap(~Method*Replicate, ncol = 3, scales = "free")
```
::::