---
title: "Single-cell Visualizations"
subtitle: "Workshop on Advanced Data Visualization"
author: "Lokesh Mano"
format: 
    html:
        font-size: 10
code-annotations: hover
---

The visualization tutorila here is inspired from the [**NBIS workshop on Single-cell RNA-seq Analysis**](https://nbisweden.github.io/workshop-scRNAseq/). The visualization tutorial her eare part of the single-cell workshop as well.

# Get data

In this tutorial, we will run all tutorials with a set of 8 PBMC 10x datasets from 4 covid-19 patients and 4 healthy controls, the samples have been subsampled to 1500 cells per sample. We can start by defining our paths.

```{r, eval = FALSE, warning=FALSE, message=FALSE}
path_covid <- "./data/covid/"
if (!dir.exists(path_covid)) dir.create(path_covid, recursive = T)

download.file("https://lu.box.com/shared/static/205zwfegtinm49yuelbvxawln6r3wyjd", destfile = file.path(path_covid, "seurat_covid_raw.rds"))
download.file("https://lu.box.com/shared/static/nf5eps2uus9qcjx0ml0n313p9a8r1vb7", destfile = file.path(path_covid, "seurat_covid_qc_dr_clst.rds"))

# If downloading via R terminal fails, you can download the images directly using the following links:
# seurat_covid_raw.rds: https://lu.box.com/s/205zwfegtinm49yuelbvxawln6r3wyjd
# seurat_covid_qc_dr_clst.rds: https://lu.box.com/s/nf5eps2uus9qcjx0ml0n313p9a8r1vb7
```

With data in place, now we can start loading libraries we will use in this tutorial.

```{r, warning=FALSE, message=FALSE}
suppressPackageStartupMessages({
    library(Seurat)
    library(Matrix)
    library(ggplot2)
    library(patchwork)
})
```

