---
title: "Dynamic plotting"
subtitle: "Quarto documents containing dynamic applications"
author: "Katja Kozjek"
institute: "NBIS, SciLifeLab"
format: 
  html:
    toc: true
    toc-depth: 2
    number-sections: true
---

::: {.callout-note}
In these exercises you will learn how to embed dynamic components, such as interactive plots and widgets, directly into a Quarto HTML document. 

By the end of this exercise, you will be able to: 

- Embed dynamic plots directly into Quarto documents.
- Discover and apply different techniques for creating and improving dynamic visualizations.
:::

# Instructions

- Open new Quarto document. 
- Create first chunk where you will load necessary R libraries. 
- Copy the following into it: 

````{.r}
```{{r}}
#| warning: false
#| message: false
library(palmerpenguins)
library(leaflet)
library(plotly)
library(ggiraph)
library(crosstalk)
library(DT)
library(ggplot2)
```
````

# Dynamic maps with leaflet 
The `leaflet` package allows creating dynamic and interactive maps using the Leaflet JavaScript library. The main advantage of using leaflet is its flexibility and that using leaflet in R is really easy.

Let's have a look into simple example: 

````{.r}
```{{r}}
leaflet() %>%
  addTiles() %>%  # Add default OpenStreetMap map tiles
  addMarkers(lng=13.1870, lat=55.7074, popup="Lund")
```
````
- Start with the `leaflet()` function. 
- New layers are added with the pipe operator (`%>%`). 
- With `addTiles()` the default basemap is added.
- To add different types of markers to the maps `addMarkers()`, `addCircleMarkers()`, `addAwesomeMarkers()` can be used.
- `popup` argument are used to display important information about a point, they appear when you click over a marker.

The example above results in: 
```{r}
#| echo: false
library(leaflet)
leaflet() %>%
  addTiles() %>%  # Add default OpenStreetMap map tiles
  addMarkers(lng=13.1870, lat=55.7074, popup="Lund")
```

## Exercise:

In the `palmerpenguins` data, penguins are found on three islands: Biscoe, Dream and Torgersen. 

``` markdown
island_coordinates <- data.frame(lng = c(-66.671, -64.225, -64.074100),
                                lat = c(-66.174, -64.725, -64.772694))
```

- Your goal is to generate dynamic map using the leaflet. 

**Advice:** Start building dynamic map step-by-step, run code after each step and pay attention to what is happening. To get help with some functions, click [here](https://r-charts.com/spatial/interactive-maps-leaflet/). 

- Make a marker for three islands.
- Add popup information (Biscoe, Dream and Torgersen island).
- Use `setView()` function to set a center point and a zoom level.
- Change the marker style to circle markers, and add color.
- Replace `addTiles()` function with `addProviderTiles(providers$Esri.WorldImagery)`, which allows to visualize the map with real images.

# plotly

We can create a `plotly` plot using the function `plot_ly()`. 
Typical composition of a plotly plot:

`plot_name <- plot_ly(data = ..., x = ~ ..., y = ~ ...)`

It is also possible to build ggplot object first and then transform it to plotly plot using the `ggplotly()` function. This works fairly well for simple plots, although it is usually a better option to build `plotly` plots from scratch.

For example, penguin body mass vs. flipper length, using `ggplot()` and transformation to plotly plot using `ggplotly()` function.

````{.r}
```{{r}}
pp1 <-
    ggplot(data = penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
    geom_point(aes(color = species), size = 2) +
    scale_color_manual(values = c("darkorange","darkorchid","cyan4")) + 
    theme_minimal()

ggplotly(pp1)
```
````

```{r}
#| echo: false
#| warning: false
library(plotly)
library(palmerpenguins)

pp1 <-
    ggplot(data = penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
    geom_point(aes(color = species), size = 2) +
    scale_color_manual(values = c("darkorange","darkorchid","cyan4")) + 
    theme_minimal()

ggplotly(pp1)
```

<br>
Now, let's plot using the function `plot_ly()`. 

````{.r}
```{{r}}
plot_ly(data = penguins, x = ~flipper_length_mm, y = ~body_mass_g, 
        color = ~species, colors = c("darkorange","darkorchid","cyan4"), size=2)
```
````

```{r}
#| echo: false
#| warning: false
library(plotly)
library(palmerpenguins)

plot_ly(data = penguins, x = ~flipper_length_mm, y = ~body_mass_g, 
        color = ~species, colors = c("darkorange","darkorchid","cyan4"), size=2)
```

When running this R command may see some red `Warning messages` appear in the R Console. Often you do not have to worry about these, but if you would like to minimise them, you can add the following arguments to your `plot_ly` function.

````{.r}
```{{r}}
plot_ly(data = penguins, x = ~flipper_length_mm, y = ~body_mass_g, 
        color = ~species, colors = c("darkorange","darkorchid","cyan4"), size=2, 
        type = "scatter", mode = "markers")
```
````

We have included the additional arguments `type = ...` and `mode = ....`

- We set `type = "scatter"` to ensure our data is plotted as a scatter plot.
- We set `mode = "markers"` to ensure that each of our data points is plotted individually.

These additional arguments are often helpful, as sometimes we like to have a little more control over how our data is presented.
You will notice however that if these commands are omitted from your function, R will just work out what it thinks is the optimal presentation format (hence the warning messages informing us which options R has selected, since some details have not been user-specified).


`plot_ly` allows us to plot different plot types, for example, histograms, boxplots, scatterplots... using argument `type = ~ ...`. Example below shows how to obtain boxplot and scatterplot and how to combine these two plots together using the `subplot` command.

````{.r}
```{{r}}
p1 <- plot_ly(data = penguins, x = ~island, y = ~flipper_length_mm, 
        color = ~island, type = "box", colors = c("#10a53dFF","#541352FF","#ffcf20FF"), size=2)

p2 <- plot_ly(data = penguins, x = ~flipper_length_mm, y = ~body_mass_g, 
        color = ~species, colors = c("darkorange","darkorchid","cyan4"), size=2)

fig <- subplot(p1, p2, nrows = 2, margin = 0.1)
fig %>% 
    layout(title = "Palmer Penguin Data",
    xaxis = list(title = "Species"), 
    yaxis = list(title = "Body mass [g]"),
    xaxis2 = list(title = "Body mass [g]"), 
    yaxis2 = list(title = "Flipper length [mm]"))
```
````

Note that here:

- We are using the `subplot` command to plot the p1 and p2 plots together.
- The `nrows = 2` argument tells R to produce these plots in 2 rows.
- The `margin = 0.1` argument tells R to leave a small margin between the two plots.
- The subsequent lines (`layout()`) of code are used to add a title to our plot, and add axes labels to the plots - note that we use `xaxis` to define the x-axis label for the first plot, and `xaxis2` to define the x-axis label for the second plot (and similarly for the y-axes).

```{r}
#| echo: false
#| warning: false

library(plotly)
library(palmerpenguins)

p1 <- plot_ly(data = penguins, x = ~island, y = ~flipper_length_mm, 
        color = ~island, type = "box", colors = c("#10a53dFF","#541352FF","#ffcf20FF"), size=2)

p2 <- plot_ly(data = penguins, x = ~flipper_length_mm, y = ~body_mass_g, 
        color = ~species, colors = c("darkorange","darkorchid","cyan4"), size=2)

fig <- subplot(p1, p2, nrows = 2, margin = 0.05)
fig %>% 
    layout(title = "Palmer Penguin Data",
    xaxis = list(title = "Species"), 
    yaxis = list(title = "Body mass [g]"),
    xaxis2 = list(title = "Body mass [g]"), 
    yaxis2 = list(title = "Flipper length [mm]"))
```

## Exercise:

- In this exercise you will use [msleep](https://ggplot2.tidyverse.org/reference/msleep.html) dataset that is part of `ggplot2()` package. 
- First check the dataset and familiarise yourself with data structure.
- Explore the relationship between REM sleep and total sleep across different dietary categories of mammals. To visualise this, use a scatter plot within `plot_ly` and test different `mode` arguments.

- Next, plot the number of mammalian in each diet (vore variable) group using the `plot_ly()` and `add_histogram()` functions.
- Then, generate boxplot using the `plot_ly()` function where you show total amount of sleep and diet. Add color to each boxplot.
- Combine both plots using the `subplot` function. Polish the combined figure, i.e. label axis, adjust box color, adjust legend... 

# ggiraph

This package is a htmlwidget and a ggplot2 extension that allows you to create dynamic graphs. Any ggplot graphic can be turned into a ggiraph graphic by calling `girafe()` function.

The first step is easy. Create a ggplot just like you normally would.

:::{.callout-note}
To run the following example, you need to download the data `bcg-immunization-coverage-for-tb-among-1-year-olds.csv`. Unfortunately, this was not part of the `data.zip` you received in the beginning of the course. You can download this file [here](https://github.com/NBISweden/workshop-data-visualization-r/raw/main/topics/dynamic_plotting/data/bcg-immunization-coverage-for-tb-among-1-year-olds.csv) 
:::

````{.r}
```{{r}}
tb_data <- read.csv("./data/bcg-immunization-coverage-for-tb-among-1-year-olds.csv", sep = ",", header = T)
tb_data_filt <- tb_data[grep("^B", tb_data$Entity), ]

ggplot(tb_data_filt, aes(x = Year, y = Share_of_newborns, color = Entity)) +
    geom_line(linewidth = 1.5) +
    scale_color_manual(values = c("#004488", "#4477AA", "#EE6677", "#228833", "#CCBB44", "#997700", "#CC3311", "#009988", "#66CCEE", "#AA3377", "#BBBBBB", "#6699CC", "#CCDDAA")) +
    theme_bw() +
    labs(title = "BCG Immunization Coverage",
             x = "Year",
             y = "Share of Newborns Immunized (%)",
             color = "Country")
```
````

```{r}
tb_data <- read.csv("./data/bcg-immunization-coverage-for-tb-among-1-year-olds.csv", sep = ",", header = T)
tb_data_filt <- tb_data[grep("^B", tb_data$Entity), ]

library(ggiraph)
library(ggplot2)

ggplot(tb_data_filt, aes(x = Year, y = Share_of_newborns, color = Entity)) +
    geom_line(linewidth = 1.5) +
    scale_color_manual(values = c("#004488", "#4477AA", "#EE6677", "#228833", "#CCBB44", "#997700", "#CC3311", "#009988", "#66CCEE", "#AA3377", "#BBBBBB", "#6699CC", "#CCDDAA")) +
    theme_bw() +
    labs(title = "BCG Immunization Coverage",
             x = "Year",
             y = "Share of Newborns Immunized (%)",
             color = "Country")
```

Next, you need to decide what parts become dynamic. Here we want to focus on the lines. Therefore, we need to make `geom_line()` dynamic. We do this by loading `ggiraph` and changing `geom_line()` to `geom_line_interactive()` and adding one of the following aesthetics in each interactive layer: `tooltip`, `data_id` or `onclick`.

- `tooltip`: column of dataset that contains tooltips to be displayed when mouse is over elements.
- `data_id`: column of dataset that contains id to be associated with elements. This aesthetic is mandatory when you want to use an hover effect or when you want to enable selection of points in shiny applications.
- `onclick`: column of dataset that contains javascript function to be executed when elements are clicked.

In our case we are going to use `tooltip = Entity`. We save the plot into a variable instead of generating output from the code. Finally, we pass our new variable, **gg** to `girafe()`. You need to call `girafe(ggobj = gg)` and not `girafe(gg)`.

````{.r}
```{{r}}
gg <- ggplot(tb_data_filt, aes(x = Year, y = Share_of_newborns, color = Entity)) +
    geom_line_interactive(aes(tooltip = Entity), linewidth = 1.5) +
    scale_color_manual(values = c("#004488", "#4477AA", "#EE6677", "#228833", "#CCBB44", "#997700", "#CC3311", "#009988", "#66CCEE", "#AA3377", "#BBBBBB", "#6699CC", "#CCDDAA")) +
    theme_bw() +
    labs(title = "BCG Immunization Coverage",
             x = "Year",
             y = "Share of Newborns Immunized (%)",
             color = "Country")

girafe(ggobj = gg)
```
````

```{r}
#| echo: false
tb_data <- read.csv("./data/bcg-immunization-coverage-for-tb-among-1-year-olds.csv", sep = ",", header = T)
tb_data_filt <- tb_data[grep("^B", tb_data$Entity), ]

library(ggiraph)
library(ggplot2)

gg <- ggplot(tb_data_filt, aes(x = Year, y = Share_of_newborns, color = Entity)) +
    geom_line_interactive(aes(tooltip = Entity), linewidth = 1.5) +
    scale_color_manual(values = c("#004488", "#4477AA", "#EE6677", "#228833", "#CCBB44", "#997700", "#CC3311", "#009988", "#66CCEE", "#AA3377", "#BBBBBB", "#6699CC", "#CCDDAA")) +
    theme_bw() +
    labs(title = "BCG Immunization Coverage",
             x = "Year",
             y = "Share of Newborns Immunized (%)",
             color = "Country")

girafe(ggobj = gg)
```

With the same logic as described above we can create all kinds of plots. For example, for boxplots the key function is `geom_boxplot_interactive()`. 

## Exercise:

- You will now try to use `geom_boxplot_interactive()` function using the `msleep` dataset you already worked with.
- Create a boxplot to visualize the total amount of total sleep across different diets, similar to the plotly example above. Enhance the plot's appearance by applying techniques and options covered in the ggplot tutorial, such as custom color palettes, axis labels, and themes.

# DataTables (DT)

The DataTables (DT) package provides a quick way to make data tables dynamic. It enables tables to be searchable, sortable, and pageable automatically. The main function in this package is `datatable()`. It creates a htmlwidget to display R data objects with DataTables.

Table below displays the first 5 rows of the `msleep` data as rendered by DT. Note the search box and clickable sorting arrows.

````{.r}
```{{r}}
datatable(msleep, options = list(pageLength = 5))
```
````

```{r}
#| echo: false
library(ggplot2)
library(DT)

datatable(msleep, options = list(pageLength = 5))
```

<br>
Another helpful tool you can use with DT is the ordering. Using the arrow widgets for each column it is easy to order a selected column in ascending or descending order. We can use the `order` option to specify how we want to order. For example, we sort the table by columns 6 (ascending) and 9 (descending):

````{.r}
```{{r}}
datatable(msleep, options = list(
  order = list(list(6, 'asc'), list(9, 'desc'))))
```
````

```{r}
#| echo: false
library(ggplot2)
library(DT)

datatable(msleep, options = list(
  order = list(list(6, 'asc'), list(9, 'desc'))))
```

<br>
To describe what the table displays the `caption` argument is used. Placing a caption directly in the table provides helpful information for the user.

````{.r}
```{{r}}
datatable(msleep, caption="Table 1: Mammals sleep dataset")
```
````

```{r}
#| echo: false
library(ggplot2)
library(DT)

datatable(msleep, caption = "Table 1: Mammals sleep dataset")
```
<br>

We can also apply CSS styles to the table cells in a column according to the values of the cells using the function `formatStyle()`. Here, a few commonly used CSS properties as the arguments of `formatStyle()`, such as `color` and `fontWeight` are used.

````{.r}
```{{r}}
datatable(msleep, options = list(pageLength = 5), caption="Table 1: Mammals sleep dataset") %>%
  formatStyle('genus',  color = 'black', backgroundColor = 'orange', fontWeight = 'bold')
```
````

```{r}
#| echo: false
library(ggplot2)
library(DT)

datatable(msleep, options = list(pageLength = 5), caption="Table 1: Mammals sleep dataset") %>%
  formatStyle('genus',  color = 'black', backgroundColor = 'orange', fontWeight = 'bold')
```

<br>
In the example above, all styles are applied to all cells in a column unconditionally: the font color is black, the background color is orange, and the font weight is bold. That may not be useful in practice. DT has provided a few helper functions to apply conditional styles to cells, such as `styleInterval()`, `styleEqual()`, and `styleColorBar()`.

````{.r}
```{{r}}
datatable(msleep) %>% 
  formatStyle('vore', fontWeight = styleEqual('omni', c('bold'))) %>%
  formatStyle('sleep_total', color = styleInterval(c(11, 18), c('white', 'blue', 'green')),
    backgroundColor = styleInterval(11, c('gray', 'orange'))
  ) %>% 
   formatStyle(
    'sleep_rem',
    background = styleColorBar(msleep$sleep_rem, 'steelblue'),
    backgroundSize = '100% 90%',
    backgroundRepeat = 'no-repeat',
    backgroundPosition = 'center'
  ) 
```
````

```{r}
#| echo: false
library(ggplot2)
library(DT)

datatable(msleep) %>% 
  formatStyle('vore', fontWeight = styleEqual('omni', c('bold'))) %>%
  formatStyle('sleep_total', color = styleInterval(c(11, 18), c('white', 'blue', 'green')),
    backgroundColor = styleInterval(11, c('gray', 'orange'))) %>% 
   formatStyle(
    'sleep_rem',
    background = styleColorBar(msleep$sleep_rem, 'steelblue'),
    backgroundSize = '100% 90%',
    backgroundRepeat = 'no-repeat',
    backgroundPosition = 'center') 
```

For the column `vore`, the font weight in cells of which the values equal to `omni` will be bold. For `sleep_total`, both foreground and background colors are used: white for values smaller than 11, blue for those between 11 and 18, and green for those greater than 18; gray background for values below 11, and orange for those above 11. For `sleep_rem`, a bar graph is presented as the background. 

## Exercise:

- To test DT package  you will use files `counts_raw.txt` and `metadata.csv`. 
- Read-in these two files, if not sure how, go back to the material from Day 1. 
- Displays the first 10 rows of `counts_raw.txt` and `metadata.csv`. Once tables are rendered and displayed play around with the 
search box and clickable sorting arrows.
- Add captions to both tables. 
- In the `metadata.csv` order column `Replicate` by descending order.
- Finally, in the `counts_raw.txt` for `Sample_1` background use bar graph using `styleColorBar()`. For `Sample_3` use `styleInterval()` and color foreground and background, decide on the interval size yourself.

# crosstalk

The `crosstalk` package can incorporate additional dynamic functionality in a Quarto document. As the name implies, it allows linking 
multiple (crosstalk-compatible) `htmlwidgets`. This allows functionality that looks interactive as in a Shiny application, but does not require Shiny Server.

The `crosstalk `package is powerful because it lets you create a `SharedData` object. The primary use for `SharedData` is to be passed to crosstalk-compatible widgets in place of a data frame. Each `SharedData$new(...)` call makes a new "group" of widgets that link to each other, but not to widgets in other groups.

The SharedData constructor has two optional parameters. The first is `key` and the second is `group`. Detailed explanation is found [here](https://rstudio.github.io/crosstalk/using.html).

**One dataset, two plots**
<br>

1. Create a `SharedData` object
2. Make htmlwidgets with `SharedData` input
    - Here, we do this first by making `ggplot` objects…
    - …and then converting them to `ggplotly` objects
3. Output results

For step 3, we use the `crosstalk::bscols()` function to put the resulting interactive plots in a row (similar to `grid.arrange`). In the output below, click on a point on one plot and notice that the point related to the same penguin is highlighted in the other plot.

````
```{{r}}
library(palmerpenguins)
library(crosstalk)
library(ggplot2)

# make SharedData object
penguins_db <- SharedData$new(penguins)

# make ggplots using SharedData object
gg_plot1 <- ggplot(data = penguins_db, aes(x = bill_length_mm, y = bill_depth_mm)) +
  geom_point(aes(color = species, 
                 shape = species),
             size = 2)  +
  scale_color_manual(values = c("darkorange","darkorchid","cyan4"))

gg_plot2 <- ggplot(data = penguins_db, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point(aes(color = species, 
                 shape = species),
             size = 2) +
  scale_color_manual(values = c("darkorange","darkorchid","cyan4"))

# convert ggplots to ggplotly
plotly_plot1 <- ggplotly(gg_plot1)
plotly_plot2 <- ggplotly(gg_plot2)

# compose output
bscols(plotly_plot1, plotly_plot2)
```
````

```{r}
#| echo: false
library(palmerpenguins)
library(crosstalk)
library(ggplot2)

# make SharedData object
penguins_db <- SharedData$new(penguins)

# make ggplots using SharedData object
gg_plot1 <- ggplot(data = penguins_db, aes(x = bill_length_mm, y = bill_depth_mm)) +
  geom_point(aes(color = species, 
                 shape = species),
             size = 2)  +
  scale_color_manual(values = c("darkorange","darkorchid","cyan4"))

gg_plot2 <- ggplot(data = penguins_db, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point(aes(color = species, 
                 shape = species),
             size = 2) +
  scale_color_manual(values = c("darkorange","darkorchid","cyan4"))

# convert ggplots to ggplotly
plotly_plot1 <- ggplotly(gg_plot1)
plotly_plot2 <- ggplotly(gg_plot2)

# compose output
bscols(plotly_plot1, plotly_plot2)
```
<br>

**One dataset, two widgets**
<br>

Crosstalk is not limited to plots; it works with other `htmlwidgets`, such as DT tables and leaflet maps. To use it, simply pass a `SharedData` object into the functions of these packages in the same way you would a standard dataframe.

In the following example, we will now make a plot interact with a table instead of another plot. 

1. Create a `SharedData` object
2. Make `htmlwidgets` with `SharedData` input
    - Before this was two `ggplotly`
    - Now, we have one `ggplotly` and one `datatable`
3. Output results

````
```{{r}}
library(palmerpenguins)
library(crosstalk)
library(ggplot2)
library(DT)

# make SharedData object
penguins_db <- SharedData$new(penguins)

# make ggplots using SharedData object
gg_plot1 <- ggplot(data = penguins_db, aes(x = bill_length_mm, y = bill_depth_mm)) +
  geom_point(aes(color = species, 
                 shape = species),
             size = 2)  +
  scale_color_manual(values = c("darkorange","darkorchid","cyan4"))

# make htmlwidgets
plotly_plot1 <- ggplotly(gg_plot1)
dt_penguins <- datatable(penguins_db)

# compose output
bscols(plotly_plot1, dt_penguins)
```
````

```{r}
#| echo: false
library(palmerpenguins)
library(crosstalk)
library(ggplot2)
library(DT)

# make SharedData object
penguins_db <- SharedData$new(penguins)

# make ggplots using SharedData object
gg_plot1 <- ggplot(data = penguins_db, aes(x = bill_length_mm, y = bill_depth_mm)) +
  geom_point(aes(color = species, 
                 shape = species),
             size = 2)  +
  scale_color_manual(values = c("darkorange","darkorchid","cyan4"))

# make htmlwidgets
plotly_plot1 <- ggplotly(gg_plot1)
dt_penguins <- datatable(penguins_db)

# compose output
bscols(plotly_plot1, dt_penguins)
```
<br>

**Multiple datasets: Two datasets, two widgets**
<br>
To link multiple `SharedData` objects, we can specify a group argument when constructing the `SharedData` objects. Group is a character string that servers as a unique identifier for a set of `SharedData` objects across whom the key information should be transmitted.

1. Create a `SharedData` objects
2. Make `htmlwidgets` with `SharedData` input
3. Output results

:::{.callout-important}
To run the following example, you need to download the data `dream_location_filt.csv`. Unfortunately, this was not part of the `data.zip` you recieved in the beginning of the course. You can download this file [here](https://github.com/NBISweden/workshop-data-visualization-r/raw/main/topics/dynamic_plotting/data/dream_location_filt.csv) 
:::

````
```{{r}}
library(dplyr)
library(leaflet)
library(palmerpenguins)
library(crosstalk)

dream_location <- read.csv("./data/dream_location_filt.csv", sep = ",", header = T)
dream_penguins <- penguins %>% filter(island == "Dream")

# make SharedData object
penguins_db <- SharedData$new(dream_penguins, group = "location")
locations_db <- SharedData$new(dream_location, group = "location")

# make htmlwidgets
dt_penguins <- datatable(penguins_db)
lf_stations <- leaflet(locations_db) %>% addTiles() %>% addMarkers()

# compose output
bscols(dt_penguins, lf_stations, widths = c(12, 12))
```
````

```{r}
#| echo: false
library(dplyr)
library(leaflet)
library(palmerpenguins)
library(crosstalk)

dream_location <- read.csv("./data/dream_location_filt.csv", sep = ",", header = T)
dream_penguins <- penguins %>% filter(island == "Dream")

# make SharedData object
penguins_db <- SharedData$new(dream_penguins, group = "location")
locations_db <- SharedData$new(dream_location, group = "location")

# make htmlwidgets
dt_penguins <- datatable(penguins_db)
lf_stations <- leaflet(locations_db) %>% addTiles() %>% addMarkers()

# compose output
bscols(dt_penguins, lf_stations, widths = c(12, 12))
```

## Exercise:

- Start by creating a SharedData object using the `SharedData$new()` function from the `crosstalk` package. Use the `msleep` dataset as input.
- Create a scatter plot using `ggplot2` with the SharedData object. Map `sleep_rem` to the x-axis, `sleep_total` to the y-axis, and `vore` to the color aesthetic.
- Convert the ggplot object to obtain plotly plot.
- Use `crosstalk` filters to add dynamic applications. Add a checkbox filter for the `vore` variable, a slider filter for `sleep_total`, and a select filter for `conservation`.
- Combine the filters and the plotly plot into a single layout using the `bscols()` function. Specify the widths for the filters and the plot.

# ObservableJS (OJS)

ObservableJS is a relatively new approach that also allows dynamic features to be included in a Quarto document. It is an entirely separate language outside of R that uses JavaScript and allows excellent functionality similar what is provided by a Shiny Server.

OJS works in any Quarto document (plain markdown as well as Jupyter and Knitr documents). Just include your code in an `{ojs}` executable code block. Similar to what you learned in Quarto lab, there are many options available to customize the behavior of `{ojs}` [code cells](https://quarto.org/docs/reference/cells/cells-ojs.html), including showing, hiding, and collapsing code as well as controlling the visibility and layout of outputs. Cell options are specified with `//|`. 

The most important cell option to be aware of is the `echo` option, which controls whether source code is displayed. In the code below `echo: false`, meaning source code is not diplayed in the final rendered report.

````
```{{ojs}}
//| echo: false
data = FileAttachment("./data/palmer-penguins.csv").csv({ typed: true })
```
````

Let's have a look into a simple example using Palmer Penguins dataset. Here we look at how penguin body mass varies across both sex and species (the provided inputs to filter the dataset by bill length and island are used).

```{ojs}
//| echo: false
data = FileAttachment("./data/palmer-penguins.csv").csv({ typed: true })

viewof bill_length_min = Inputs.range(
  [32, 50], 
  {value: 35, step: 1, label: "Bill length (min):"}
)
viewof islands = Inputs.checkbox(
  ["Torgersen", "Biscoe", "Dream"], 
  { value: ["Torgersen", "Biscoe"], 
    label: "Islands:"
  }
)

filtered = data.filter(function(penguin) {
  return bill_length_min < penguin.bill_length_mm &&
         islands.includes(penguin.island);
})

Plot.rectY(filtered, 
  Plot.binX(
    {y: "count"}, 
    {x: "body_mass_g", fill: "species", thresholds: 20}
  ))
  .plot({
    facet: {
      data: filtered,
      x: "sex",
      y: "species",
      marginRight: 80
    },
    marks: [
      Plot.frame(),
    ]
  }
)
```

To begin, we create an `{ojs}` cell to load data from a CSV file (e.g., palmer-penguins.csv) using a [FileAttachment](https://observablehq.com/documentation/data/files/file-attachments). File attachments support various formats such as CSV, TSV, and JSON, making them a convenient way to load pre-prepared datasets for analysis.

To get pre-prepared dataset `palmer-penguins.csv` load `palmerpenguins` library, save `penguins` data as `.csv` file and then load this data as described above. 

````
```{{ojs}}
data = FileAttachment("./data/palmer-penguins.csv").csv({ typed: true })
```
````

Now we loaded the external CSV file into the Observable environment.
<br>

**CSV files:**
<br>

- The `.csv()` method parses the CSV file into a JavaScript array of objects, where each row in the CSV becomes an object, and the column headers become the keys.
- The `{ typed: true }` option ensures that the data types of the columns are automatically inferred. For example:
  - Numeric columns are converted to numbers.
  - Text columns remain as strings.
  - Dates are parsed into JavaScript `Date` objects (if applicable).

::: {.callout-note title="Data Sources - R"}
The data you want to use with OJS might not always be available in raw form. Often you will need to read and preprocess the raw data using R. You can perform this preprocessing during document render (in an `{r}` code cell) and then make it available to `{ojs}` cells via the `ojs_define()` function.

Here is an example. Read the data into R, do some grouping and summarization, then make it available to OJS using `ojs_define`:

````
```{{r}}
library(readr)
library(dplyr)

penguins_filt = read_csv("palmer-penguins.csv") %>%
  group_by(species) %>%
  summarise(mean_body_mass = mean(body_mass_g, na.rm = TRUE))

ojs_define(penguinsdata = penguins_filt)
```
````

Once JS object is created, it needs to be transposed using the `transpose()` function. This function will convert column-oriented datasets (like the ones used in R) into the row-oriented datasets used by many JavaScript plotting libraries.

````
```{{ojs}}
ojsdata <- transpose(penguinsdata)
```
````
:::

The example plotted (Palmer Penguins dataset) above does not display the entire dataset but instead a filtered subset. To create this filter, we need inputs and the ability to use their values in our filtering function. This is achieved using the `viewof` keyword along with standard [Inputs](https://observablehq.com/documentation/inputs/overview).

````
```{{ojs}}
viewof bill_length_min = Inputs.range(
  [32, 50], 
  {value: 35, step: 1, label: "Bill length (min):"}
)
viewof islands = Inputs.checkbox(
  ["Torgersen", "Biscoe", "Dream"], 
  { value: ["Torgersen", "Biscoe"], 
    label: "Islands:"
  }
)
```
````

The range input above (created with `Inputs.range()`) specifies the total range of values a user can choose from (32 to 50), along with options for:

- `value`: the default starting value or option of an input (here, 35)
- `step`: the increments along the range (here, change by increments of 1)
- `label`: a label informing or prompting the user (here, Bill length (min))

The checkbox input above (created with `Inputs.checkbox()`) allows the user to choose any of a given set of values, in this case three islands where penguins are found. 

- `value`: the default starting values or options of an input (here, Torgersen, Biscoe)
- `label`: a label informing or prompting the user (here, Islands)

The input values (here bill_length_min and islands) are generated and can be referenced in any cell, and the cell will run whenever the input changes.

Next step is to write the filtering function that will transform the data read from the CSV using the values of `bill_length_mm` and `island`.

````
```{{ojs}}
filtered = data.filter(function(penguin) {
  return bill_length_min < penguin.bill_length_mm &&
         islands.includes(penguin.island);
})
```
````

The `filtered` variable should contains an array of penguin objects that meet both conditions: `bill_length_mm` is greather than `bill_length_min` and `island` is one of the values in the `islands`. 

Finally, let's plot the filtered data using [Observable Plot](https://observablehq.com/plot/) (an open-source JavaScript library for quick visualization of tabular data).

````
```{{ojs}}
Plot.rectY(filtered, 
  Plot.binX(
    {y: "count"}, 
    {x: "body_mass_g", fill: "species", thresholds: 20}
  ))
  .plot({
    facet: {
      data: filtered,
      x: "sex",
      y: "species",
      marginRight: 80
    },
    marks: [
      Plot.frame(),
    ]
  }
)
```
````

Observe that the `filtered` variable is referenced directly in the plotting code without any special syntax. This ensures that the plot is automatically updated whenever the `filtered` variable changes, which in turn is triggered by any updates to the input (bill_length_min, islands) values.

## Exercise:

For this exercise is recommended to create a new Quarto (`.qmd`) notebook. You will work with `msleep` dataset, and your goal is to 
generate a scatter plot that shows the relationship between mammalian REM and total sleep where diet (`vore` variable) is used as the input to filter the dataset.


- Prepare data in R and pass it to OJS using `ojs_define`. Transpose JS object.
- Display the raw dataset in a table using `Inputs.table` for better data exploration.
- Create input using the `Inputs.checkbox` for `vore` variable, to allow dynamic filtering of the dataset.
- Filter the dataset based on user input using JavaScript's `filter` function.
- Visualize the filtered data using Observable Plot. Use `Plot.dot()` to create scatterplot.